The Galactic Midplane Is Not a Plane: Implications for Dynamical Analysis with \textit{Gaia} Data and Beyond

### 2019-05-10 paper/ms.tex
Add citation to Buist paper

Now suppose this observer makes a measurement of the star's position and
velocity at the teal point or the orange point (i.e. at two different orbital
phases). Then the teal and orange curves correspond to the orbits that the
observer would compute for each point based on the potential model with the
offset midplane. In action space, this would correspond to a different value
of $J_z$ for the teal and orange points. In this way, assuming the wrong
coordinate system induces a phase-dependence in the actions estimated for the
star, which in the correct potential (in this example, the one with the
correct midplane) should be phase-independent. This example uses an offset in $z$, but one would expect an analogous effect from an offset in another coordinate, such as the distance to the Galactic center or the local standard of rest.

@@

\item To demonstrate the previous point, we measured the local galactic
midplane along the Solar circle in three different high-resolution, zoom-in
simulations of Milky Way mass galaxies from the FIRE collaboration, as well as a controlled simulation of the interaction of the Milky Way with Sagittarius. We found
that the midplane varies as a function of azimuth at the Solar circle by
$\sim$185, 162, 84 pc (middle $90\%$) for the three cosmological simulations we considered
(\mi{}, \mf{}, and \mm{}, respectively) and by $\sim$30 and $\sim$100 pc before and after the interaction in the controlled simulation. Comparison of the corresponding velocity variations with recent measurements in the Milky Way suggests that these values are consistent with expectations for midplane variations in the Milky Way itself, which have not yet been measured as a function of azimuth.

@@

\item This work underlines the importance of combining chemistry and dynamics.
Since chemical tagging \citep{2002ARA&A..40..487F} is not subject to the same systematic errors discussed in this paper, it should be used to confirm dynamical associations and to offset the effect of these systematic errors on the action-space distribution function. %The importance of combining chemistry and dynamics for Galactic archaeology was recently pointed out by \citet{2019arXiv190210719K}.

@@

\acknowledgments
We would like to thank Megan Bedell, Robert A. Benjamin, Tobias Buck, Elena
D'Onghia, Benoit Famaey, and Adrian Price-Whelan for helpful discussions. A.B.
would like to thank Todd Phillips for helpful discussions. A.B. was supported
in part by the Roy \& Diana Vagelos Program in the Molecular Life Sciences and
the Roy \& Diana Vagelos Challenge Award. R.E.S's, K.V.J.'s and C.F.P.L.'s work was
performed in part during the Gaia19 workshop and the 2019 Santa Barbara Gaia
Sprint (also supported by the Heising-Simons Foundation), both hosted by the
Kavli Institute for Theoretical Physics at the University of California, Santa
Barbara. This work used the Extreme Science and Engineering Discovery
Environment (XSEDE), which is supported by National Science Foundation grant
number OCI-1053575. K.V.J.'s contributions were supported in part by the
National Science Foundation under grants NSF PHY-1748958 and NSF AST-1614743.
M.-M.M.L. was partly supported by NSF grant AST18-15461. The work of D.A.-A.,
A.B., D.W.H., R.E.S., K.V.J., M.-M.M.L., and M.K.N. is supported by the Simons
Foundation.

>>>>

Now suppose this observer makes a measurement of the star's position and
velocity at the teal point or the orange point (i.e. at two different orbital
phases). Then the teal and orange curves correspond to the orbits that the
observer would compute for each point based on the potential model with the
offset midplane. In action space, this would correspond to a different value
of $J_z$ for the teal and orange points. In this way, assuming the wrong
coordinate system induces a phase-dependence in the actions estimated for the
star, which in the correct potential (in this example, the one with the
correct midplane) should be phase-independent.

This example uses an offset in $z$, but one would expect an analogous effect
from an offset in another coordinate, such as the distance to the Galactic
center or the local standard of rest. A similar effect in which actions gain
time-dependence due to a time-varying potential was pointed out by
\citet{2015A&A...584A.120B}.

@@

\item To demonstrate the previous point, we measured the local galactic
midplane along the Solar circle in three different high-resolution, zoom-in
simulations of Milky Way mass galaxies from the FIRE collaboration, as well as a controlled simulation of the interaction of the Milky Way with Sagittarius. We found
that the midplane varies as a function of azimuth at the Solar circle by
$\sim185$, $162$, $84\,\pc$ (middle $90\%$) for the three cosmological simulations we considered
(\mi{}, \mf{}, and \mm{}, respectively) and by $\sim30$ and $\sim100\,\pc$ before and after the interaction in the controlled simulation. Comparison of the corresponding velocity variations with recent measurements in the Milky Way suggests that these values are consistent with expectations for midplane variations in the Milky Way itself, which have not yet been measured as a function of azimuth.

@@

\item This work underlines the importance of combining chemistry and dynamics.
Since chemical tagging \citep{2002ARA&A..40..487F} is not subject to the same systematic errors discussed in this paper, it should be used to confirm dynamical associations and to offset the effect of these systematic errors on the action-space distribution function.

@@

\acknowledgments
We would like to thank Megan Bedell, Robert A. Benjamin, Tobias Buck, Elena
D'Onghia, Benoit Famaey, and Adrian Price-Whelan for helpful discussions. A.B.
would like to thank Todd Phillips for helpful discussions. A.B. was supported
in part by the Roy \& Diana Vagelos Program in the Molecular Life Sciences and
the Roy \& Diana Vagelos Challenge Award. K.V.J.'s, C.F.P.L.'s, and R.E.S.'s work was
performed in part during the Gaia19 workshop and the 2019 Santa Barbara Gaia
Sprint (also supported by the Heising-Simons Foundation), both hosted by the
Kavli Institute for Theoretical Physics at the University of California, Santa
Barbara. This work used the Extreme Science and Engineering Discovery
Environment (XSEDE), which is supported by National Science Foundation grant
number OCI-1053575. K.V.J.'s contributions were supported in part by the
National Science Foundation under grants NSF PHY-1748958 and NSF AST-1614743.
M.-M.M.L. was partly supported by NSF grant AST18-15461. The work of D.A.-A.,
A.B., D.W.H., R.E.S., K.V.J., M.-M.M.L., and M.K.N. is supported by the Simons
Foundation.
